Subset coercions in \Coq

### 2006-07-16 papers/article/main.tex
Update to new presentation with types in the coercion derivation judgement

\input{typing-algo}
\renewcommand{\AppAFull}[9][\Gamma]{
\TAX{App}
{$\tgen{#1}{#2}{#3} \suba \Pi #4 : #5. #6$}
{$\tgen{#1}{#8}{#9}$} % \quad \tgen{#1}{#9, #5}{#7'}$}
{$\subalgo{#1}{#9}{#5}$}
{$\tgen{#1}{(#2 #8)}{#6 [ #8 / #4 ]}$}
{}
}
\input{subtyping-algo}
\renewcommand{\SubSubAs}[1][\Gamma]{%
\UAX{SubSub}
{$\subalgoimpl{#1}{c}{U}{T}$}
{$\subalgoimpl{#1}{c[\eltpit~\ctxdot]}{\mysubset{x}{U}{P}}{T}$}
{}
}

@@

Informally, this idea is already present in mathematics. When you have a
subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, you usually prove \emph{first}
that you apply it only to elements having the desired property. 
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. We would like for example
to prove that our usage of partial functions is correct only after the
program is wrote. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and the
properties which we use to reason about them.

@@

\subsubsection{The typing judgment}
\begin{figure}[t]
  \begin{center}
    \def\infvspace{0.1in}
    \def\seq{\typed}
    \def\fCenter{\wf}

@@

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell{} typing}
  \label{fig:typing-decl-rules}
  \vspace{-1em}
\end{figure}
%
\Russell{}'s typing system (figure \vref{fig:typing-decl-rules}) is based
on the Calculus of Constructions \cite{coquandhuet88,Barras99}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is equivalent to $U$ in environment $`G$. 
Following the presentation as a pure type system \cite{barendregt91}, 
the set of sorts $\PTSsorts$ is defined as $\setproptype$. As usual, we
let $s, s_i$ for $i `: `N$ range over sorts. The
axioms are $\PTSaxioms = \{ (\Set, \Type), (\Prop, \Type) \}$ and the
set of rules $\PTSrules$ is defined by the following (functional) relation:
% \begin{table}[h]
%   \vspace{-1.5em}
  \axiomsd
%   \vspace{-1em}
%   \caption{Definition of $\PTSrules$}
%   \label{fig:PTSrules}
% \end{table}

@@

Some basic properties of this system are well-formedness of
environments: if $`G \typed t : T$ then $\wf{`G}$ and
well-sortedness: if $`G \typed t : T$ then $T = \Type$ or $`G \typed T :
s$.

@@

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
\emph{subset equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also conveys the idea that it can
include the usual $\beta\pi$-conversion directly in the judgement, 
although it may not be possible if we have cummulative universes.
The judgment $`G \typed T \subd{} U$ (figure
\vref{fig:conversion-decl-rules}  ) reads $T$ is convertible to $U$ in
environment $`G$. In the following, it is always assumed that the two
input types of the equivalence judgment are well-sorted, but maybe not by the same sort.

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. It is well-known that $\beta\pi$ is
terminating only on well typed terms, so we have to
ensure that our rules, when read from bottom to top, keep
well-typedness. That's the purpose of
the $`G \typed U : s$ premise of the \irule{SubTrans} rule. This rule
ensures that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules do context closure for dependent
products and sums. It is remarkable that we use contravariance
\cite{journals/toplas/Castagna95} for
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symetry. However, it will
become important when we create coercions (see
\vref{fig:subtyping-impl-rules}).

The really interesting rules are \irule{SubSub} and
\irule{SubProof}. The first one allows to use an object of a subset type
as an object of its support type. The later allows (maybe abusively) to
consider an object of any type as an object of any subset based on
this type. By the assumption on the input types of the judgment, we know
that the property $P$ is well-formed, but we do not care about the proof.

\subsubsection{Properties}
\begin{paragraph}{Assumptions}
  We have proved basic metatheoretical results for this system in \Coq{}
  \cite{coq/Russell/meta}. We have assumed that it is strongly normalizing (SN) and has the
  following subject reduction (SR) property: if $`G \typed t : T$ and $t
  "->"_{\beta\pi} t'$ then $`G \typed t' : T$. Gang Chen
  \cite{ChenPhD} has studied various type systems from the
  $\lambda$-cube extended by subtyping or coercive subtyping \cite{DBLP:conf/csl/Luo96},
  including the Calculus of Constructions, and proved such results as
  SN and SR for them. It requires a very careful analysis of the system 
  to avoid cycles in the proof due to the presence of dependent types and
  conversion ; hence we leave it for future work. Strong normalization
  could also be proved, perhaps more easily, using the generic proof method developed in
  \cite{geuvers95}, by finding a proper interpretation of \Russell terms
  into the Calculus of Constructions.
\end{paragraph}

As all functionnal PTS, \Russell{} has the unicity of sorting property,
hence we can prove:
\begin{lemma}[Well-sortedness of equivalence derivations]
  If $`G \typed T : s_1$, $`G \typed U : s_2$ and $`G \typed T \sub U$
  then $`G \typed T, U : s$\footnote{We write $`G \typed T, U : s$ for $`G \typed T : s$ and $`G \typed U :
  s$.} and $s = s_1 = s_2$.
\end{lemma}
\begin{proof}
  We will use the abbreviation $`G \typed T \sub U : s$ for $`G \typed T,
  U : s$ and $`G \typed T \sub U$ in the following.

  By induction on the subtyping derivation:
  \begin{list}{-}
  \item \rulename{SubConv}: We have $T \eqbp U$, we apply unicity of sorting.

  \item[-] \rulename{SubTrans}: We have by induction $`G \typed T \sub S : s$ and 
    $`G \typed S \sub U : s$. Hence we have $`G \typed T, U : s$.

  \item Other cases are straightforward.
  \end{list}
\end{proof}

The essential properties of the $\subd$ judgement are 
its reflexivity: if $`G \typed A : s$, then $`G \typed A \subd A$,
symetry: if $`G \typed A, B : s$ then $`G \typed A \subd B "<=>" `G \typed
B \subd A$ and transitivity by \irule{SubTrans}.

As typing and equivalence are mutually recursive, we will generaly prove
properties by mutual induction on the two derivations.
Our system has good structural properties like weakening, thinning and
substitution.

\begin{lemma}[Weakening]
  Suppose $\wf{`G, x : S, `D}$.
  If $`G, `D \typed t : T$ then $`G, x : S, `D \typed t : T$. 
  If $`G, `D \typed T \sub U : s$ then $`G, x : S, `D \typed T \sub U : s$.
\end{lemma}
\begin{proof}
  Straightforward mutual induction.
\end{proof}

Similarly for thining and substitution. An important property for
dependent type systems is to have invariance of judgments which are
definitionaly equal (here equivalent). Hence the following lemma:
\begin{lemma}
  Suppose $`G \typed U \sub V : s$.
  If $`G, x : V, `D \typed t : T$ then $`G, x : U, `D \typed t : T$.
  If $`G, x : V, `D \typed S \sub T : s'$ then
  $`G, x : U, `D \typed S \sub T : s'$.
\end{lemma}
\begin{proof}
  By mutual induction on the typing and equivalence derivations.
\end{proof}

We stop here about the theoretical properties of this first system, as
it gives no new insights for our purpose and we will focus more on the
algorithmic system's properties.

\subsubsection{Algorithmic typing}
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell{} algorithmic typing, new rules}
  \label{fig:typing-algo-rules}
  \vspace{-1em}
\end{figure*}

@@

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
$T `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
\TODO{App?, SumDep?, Pi?}. Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. This notion is presented in \PVS{} by
means of a $\mu_0$ operator which was readily adaptable in our setting
\ref{fig:mualgo}. In Gang Chen's work, the notion is formalized by a
new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator.
\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two well-sorted types. We denote it by $`G \typea T
  \suba U$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity rule which is not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Elimination of transitivity]
  If $`G \typea T \suba U$ and $`G \typea U \suba V$, then $`G \typea T
  \suba V$.
\end{theorem}
\begin{proof}
  By lexicographic induction on the pair of derivations.
  \TODO{example}
\end{proof}
\end{paragraph}

Classicaly, we have the following correspondence between the two
systems:
\begin{theorem}[Soundness]
  If $`G \typea t : T$ then $`G \typed t : T$.
  If $`G \typea T \suba U$ then $`G \typed T \subd U$.
\end{theorem}

\begin{theorem}[Completeness]
  If $`G \typed t : T$ then $`G \typea t : T' \suba T$.
  If $`G \typed T \subd U$ then $`G \typea T \suba U$.
\end{theorem}

Finaly, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing and equivalence]
  $`A `G~t, `E T, `G \typea t : T$ and $`A `G~T~U, `G \typea T \suba U$ are decidable
  problems.
\end{theorem}

>>>>

\input{typing-algo}
% \renewcommand{\AppAFull}[9][\Gamma]{
% \TAX{App}
% {$\tgen{#1}{#2}{#3} \suba \Pi #4 : #5. #6$}
% {$\tgen{#1}{#8}{#9}$} % \quad \tgen{#1}{#9, #5}{#7'}$}
% {$\subalgo{#1}{#9}{#5}{s}$}
% {$\tgen{#1}{(#2 #8)}{#6 [ #8 / #4 ]}$}
% {}
% }
\input{subtyping-algo}
\renewcommand{\SubSubAs}[1][\Gamma]{%
\UAX{SubSub}
{$\subalgoimpl{#1}{c}{U}{T}{\Set}$}
{$\subalgoimpl{#1}{c[\eltpit~\ctxdot]}{\mysubset{x}{U}{P}}{T}{\Set}$}
{}
}

@@

Informally, this idea is already present in mathematics. When you have a
subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, you usually prove \emph{first}
that you apply it only to elements having the desired property. 
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. We would like for example
to prove that our usage of partial functions is correct only after the
program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and the
properties which we use to reason about them.

@@

\subsubsection{Type system}
\begin{figure}[t]
  \begin{center}
    \def\infvspace{0.1in}
    \def\seq{\typed}
    \def\fCenter{\wf}

@@

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell{} typing}
  \label{fig:typing-decl-rules}
  \vspace{-1em}
\end{figure}
%
\Russell{}'s type system (figure \vref{fig:typing-decl-rules}) is based
on the Calculus of Constructions \cite{coquandhuet88,Barras99}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is equivalent to $U$ in environment $`G$. 
Following the presentation as a pure type system \cite{barendregt91}, 
the set of sorts $\PTSsorts$ is defined as $\setproptype$. As usual, we
let $s, s_i$ for $i `: `N$ range over sorts. The
axioms are $\PTSaxioms = \{ (\Set, \Type), (\Prop, \Type) \}$ and the
set of rules $\PTSrules$ is defined by the following (functional) relation:
% \begin{table}[h]
%   \vspace{-1.5em}
  \axiomsd
%   \vspace{-1em}
%   \caption{Definition of $\PTSrules$}
%   \label{fig:PTSrules}
% \end{table}

@@

@@

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
\emph{subset equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also conveys the idea that it can
include the usual $\beta\pi$-conversion directly in the judgement, 
although it may not be possible if we have cummulative universes.
The judgment $`G \typed T \subd{} U : s$ (figure
\vref{fig:conversion-decl-rules}) reads $T$ is equivalent to $U$ in
environment $`G$, both being sorted by $s$.

@@

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubSym} and
\irule{SubTrans} rules ensure that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules do context closure for dependent
products and sums. It is remarkable that we use contravariance
\cite{journals/toplas/Castagna95} for
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symetry. However, it will
become important when we create coercions (see
\vref{fig:subtyping-impl-rules}).

The really interesting rules are \irule{SubSub} and
\irule{SubProof}. The first one allows to use an object of a subset type
as an object of its support type. The later allows (maybe abusively) to
consider an object of any type as an object of any subset based on
this type. We must check that the property $P$ is well-formed, but
we do not care about its provability.

\subsubsection{Properties}
We have proved the metatheory for this system in \Coq{}
\cite{sozeau:coq/Russell/meta}. We have assumed that it is strongly
normalizing (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$. Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we prefered to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
Strong normalization could also be proved, perhaps more easily,
using the generic proof method developed in \cite{geuvers95}, by finding
a proper interpretation of \Russell terms into the Calculus of
Constructions.

So, apart from strong normalization, we have shown all the usual structural
and metatheoretic properties of a dependent programming language, like
weakening, thinning and substitution, stability by context coercion etc\ldots
We stop here on the subject of theoretical properties of this first system, as
it gives no new insights for our purpose and we will focus more on the
algorithmic system's properties. Besides, the details of the
formalization and the proof fall outside the scope of this paper.
% The essential properties of the $\subd$ judgement are 
% its reflexivity: if $`G \typed A : s$, then $`G \typed A \subd A$,
% symetry: if $`G \typed A, B : s$ then $`G \typed A \subd B "<=>" `G \typed
% B \subd A$ and transitivity by \irule{SubTrans}.

% As typing and equivalence are mutually recursive, we will generaly prove
% properties by mutual induction on the two derivations.

% \begin{lemma}[Weakening]
%   Suppose $\wf{`G, x : S, `D}$.
%   If $`G, `D \typed t : T$ then $`G, x : S, `D \typed t : T$. 
%   If $`G, `D \typed T \sub U : s$ then $`G, x : S, `D \typed T \sub U : s$.
% \end{lemma}
% \begin{proof}
%   Straightforward mutual induction.
% \end{proof}

% Similarly for thining and substitution. An important property for
% dependent type systems is to have invariance of judgments which are
% definitionaly equal (here equivalent). Hence the following lemma:
% \begin{lemma}
%   Suppose $`G \typed U \sub V : s$.
%   If $`G, x : V, `D \typed t : T$ then $`G, x : U, `D \typed t : T$.
%   If $`G, x : V, `D \typed S \sub T : s'$ then
%   $`G, x : U, `D \typed S \sub T : s'$.
% \end{lemma}
% \begin{proof}
%   By mutual induction on the typing and equivalence derivations.
% \end{proof}

\subsubsection{Algorithmic typing}

\begin{paragraph}{}
We use a standard transformation to get an algorithm from our typing
rules, integrating the conversion (or subsumption) rule
(here \irule{Coerce}) to the premises of the other rules.
We recall that $`G \type t : s$ is shorthand for $`G \type t : T$ and
$T  `= s$ with $s `: \PTSsorts$ in a algorithmic presentation.
\TODO{App?, SumDep?, Pi?}. Our algorithmic judgement $`G \typea t : T$
(figure \vref{fig:typing-algo-rules})
reads: $t$ is a well-typed term of type $T$ in environment $`G$.
We need to introduce the notion of support for subset types in order to
define our algorithmic system. Indeed, when typing an application, 
we need to ensure that the object we apply can be seen as an object of a
product type modulo the equivalence. This notion is presented in \PVS{} by
means of a $\mu_0$ operator which was readily adaptable in our setting
\ref{fig:mualgo}. In Gang Chen's work, the notion is formalized by a
new judgement $\Pi_{lub}$ which is essentialy equivalent to the operator.
\begin{figure*}[t]
  \def\infvspace{0.5em}
    \def\type{\typea}
    \def\subt{\subta}
    \def\sub{\suba}
    \def\fCenter{\typea}
    \begin{center}
%     \WfEmpty\DP

@@

%     \vspace{\infvspace}
%     \SubsetR\DP
  \end{center}
  \vspace{-2em}
  \caption{\Russell{} algorithmic typing, new rules}
  \label{fig:typing-algo-rules}
%  \vspace{-1em}
\end{figure*}

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case. We also need to eliminate the transitivity rule which is not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Elimination of transitivity]
  If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
  \suba V : s$.
\end{theorem}
\begin{proof}
  By induction on the sum of the depths of the two derivations.
  \TODO{example?}
\end{proof}
\end{paragraph}

Classicaly, we have the following correspondence between the two
systems:
\begin{theorem}[Soundness]
  If $`G \typea t : T$ then $`G \typed t : T$.
  If $`G \typea T \suba{} U : s$ then $`G \typed T \subd U : s$.
\end{theorem}

\begin{theorem}[Completeness]
  If $`G \typed t : T$ then $`G \typea t : T' \suba T$.
  If $`G \typed T \subd U : s$ then $`G \typea T \suba U : s$.
\end{theorem}

Finaly, we can state the desired property of the algorithmic system:
\begin{theorem}[Decidability of typing and equivalence]
  $`A `G~t, `E T, `G \typea t : T$ and $`A `G~T~U~s, `G \typea T \suba U :
  s$ are decidable problems.
\end{theorem}
